---
title: "Assignment 3"
author: "Insert My Name"
date: "2022-10-12"
output: html_document # you can change to other output format if you want
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Exploratory data analysis


## 1.1 (Q1)   
First, create a vector called HawksTail, the elements of which are from the Tail column of Hawks data
frame. Second, use the mean and median functions to compute the sample mean and sample median from the vector
HawksTail

```{r 1.1-1}
HawksTail <- c(Hawks$Tail)

HT_mean <- mean(HawksTail)
print("Mean:")
print(HT_mean)

HT_median <- median(HawksTail)
print("Median:")
print(HT_median)
```


## 1.2 (Q1)
Use a combination of the summarise(), mean() and median() to compute the sample mean, sample
median and trimmed sample mean (with q = 0.5) of the Hawk’s wing length and Hawk’s weight (i.e., the
Wing and Weight columns). You may need to remove the NA values. What can you say by comparing the
results of the median and the trimmed mean that you obtain

```{r 1.2-1}
Hawks %>% select(Wing,Weight) %>% summarise(Wing_avg=mean(Wing,na.rm=TRUE),Wing_med=median(Wing,na.rm=TRUE),Wing_trim_avg=mean(Wing,na.rm=TRUE,trim=0.5),Weight_avg=mean(Weight,na.rm=TRUE),Weight_med=median(Weight,na.rm=TRUE),Weight_trim_avg=mean(Weight,na.rm=TRUE,trim=0.5)) %>% print()
```



## 1.2 (Q2)
Combine them with the group_by() function to obtain a breakdown by species.

```{r 1.2-2}
Hawks %>% group_by(Species) %>% summarise(Wing_avg=mean(Wing,na.rm=TRUE),Wing_med=median(Wing,na.rm=TRUE),Wing_trim_avg=mean(Wing,na.rm=TRUE,trim=0.5),Weight_avg=mean(Weight,na.rm=TRUE),Weight_med=median(Weight,na.rm=TRUE),Weight_trim_avg=mean(Weight,na.rm=TRUE,trim=0.5)) %>% print()
```


##1.3 (Q1)
Suppose that a variable of interest X has values X1, · · · , Xn. Suppose that X1, · · · , Xn has a sample
mean A. Let a, b ∈ R be real numbers and define a new variable X˜ with X˜
1, · · · , X˜
n defined by X˜
i = aXi + b
for i = 1, 2, · · · , n. What is the sample mean of X˜
1, · · · , X˜
n as a function of a, b and A?
Now using the vector HawksTail that you created in Section 1.1 as data and letting a = 2 and b = 3, verify
your conclusion using R codes: Compute the mean of HawksTail*a+b and then compare it with the one
obtained from the mean of HawksTail and your conclusion.

```{r 1.3-1}
A <- mean(HawksTail)
a <- 2
b <- 3

HawksTail2 <- a*HawksTail+b
print(mean(HawksTail2))

print(b+a*A)
```


##1.3 (Q2)
Suppose further that X1, · · · , Xn has sample variance p and standard deviation q. What is the sample
variance of X˜
1, · · · , X˜
n? What is the sample standard deviation of X˜
1, · · · , X˜
n?
Now using the vector HawksTail that you created in Section 1.1 as data and letting a = 2 and b = 3, verify
your result using R codes again

```{r 1.3-2}
p <- var(HawksTail)
q <- sd(HawksTail)

#calculate with functions
p2 <- var(HawksTail2) %>% print()
q2 <- sd(HawksTail2) %>% print()
#calculate with formulae
print(p*a**2)
print(a*q)
```

##1.4 (Q1)
In this exercise we shall investigate the robustness of several location estimators: The sample mean, sample
median and trimmed mean.

We begin by extracting a vector called “hal” consisting of the talon lengths of all the hawks with any missing
values removed.

To investigate the effect of outliers on estimates of location we generate a new vector called “corrupted_hall”
with 10 outliers each of value 100 

We can then compute the mean of the original sample and the corrupted sample
Now let’s investigate what happens as the number of outliers changes from 0 to 1000. The code below
generates a vector called “means_vect” which gives the sample means of corrupted samples with different
numbers of outliers. More precisely, means_vect is a vector of length 1001 with the i-th entry equal to the
mean of a sample with i − 1 outliers.

Copy and modify the above code to create an additional vector called “medians_vect” of length 1001 with
the i-th entry equal to the median of a sample “corrupted_hal” with i − 1 outliers.

```{r 1.4-1}
hal <- Hawks$Hallux
hal <- hal[!is.na(hal)]

outlier_value <- 100
num_outliers <- 10
corrupted_hal <- c(hal,rep(outlier_value,times=num_outliers))

mean(hal) %>% print()
mean(corrupted_hal) %>% print()

num_outliers_vect <- seq(0,1000)
means_vect <- c()
median_vect <- c()
trim_means_vect <- c()

#how does no. of outliers affect mean
for (num_outliers in num_outliers_vect){
  corrupted_hal <- c(hal,rep(outlier_value,times=num_outliers))
  means_vect <- c(means_vect,mean(corrupted_hal))
}
#median
for (num_outliers in num_outliers_vect){
  corrupted_hal <- c(hal,rep(outlier_value,times=num_outliers))
  median_vect <- c(median_vect,median(corrupted_hal))
}
```

##1.4 (Q2)
Amend the code further to add an additional vector called “t_means_vect” of length 1001 with the i-th entry
equal to the trimmed mean of a sample with i − 1 outliers, where the trimmed mean has a trim fraction
q = 0.1

```{r 1.4-2}
#trimmed mean q=0.1
for (num_outliers in num_outliers_vect){
  corrupted_hal <- c(hal,rep(outlier_value,times=num_outliers))
  trim_means_vect <- c(trim_means_vect,mean(corrupted_hal,trim=0.1))
}
```

##1.4 (Q3)
Now you should have the vectors “num_outliers_vect”, “means_vect”, “medians_vect” and
“t_means_vect”.Now use the code below to reshape and plot the data. Recall that the function pivot_longer() below is
used to reshape the data

```{r 1.4-3}
df_means_medians <- data.frame(num_outliers=num_outliers_vect,mean=means_vect,median=median_vect,trim_mean=trim_means_vect)

means_median_plot <- df_means_medians %>% pivot_longer(!num_outliers,names_to="Estimator",values_to="Value") 
print(ggplot(means_median_plot,aes(x=num_outliers,color=Estimator,linetype=Estimator,y=Value)) + geom_line() + xlab("Number of Outliers"))

```

##1.5 (Q1)
Use the functions ggplot() and geom_boxplot() to create a box plot which summarises
the distribution of hawk weights broken down by species

```{r 1.5-1}
#box plot with weights broken down as species
weight_plot <- ggplot(data=Hawks,aes(x=Species,y=Weight)) + geom_boxplot() + xlab("Species") + ylab("Weight (g)")
print(weight_plot)
```

##1.5 (Q2)
Compute the 0.25-quantile, 0.5-quantile, 0.75-quantile of the Weight grouped by Species.
Now compare these values with the boxplot above. Can you explain which parts of the boxplot these numbers
correspond to?

```{r 1.5-2}
#compute weights grouped by species
Hawks %>% group_by(Species)%>% summarise(quantile025=quantile(Weight,probs=0.25,na.rm=TRUE),quantile050=quantile(Weight,probs=0.5,na.rm=TRUE),quantile075=quantile(Weight,probs=0.75,na.rm=TRUE)) %>% print()
```

##1.5 (Q3)
Suppose we have a sample X1, · · · , Xn. Let “q25” denote the 0.25-quantile of the sample and let “q75” denote
the 0.75-quantile of the sample. We can then define the interquartile range, denoted IQR by IQR := q75-q25.
In the context of boxplots, an outlier Xi
is any numerical value such that the following holds if either of the
following holds:
Xi < q25 − 1.5 × IQR, or
Xi > q75 + 1.5 × IQR.
Create a function called “num_outliers” which computes the number of outliers within a sample (with
missing values excluded)

```{r 1.5-3}
num_outliers <- function (val){
  val <- val[!is.na(data)]
  iq_range <- IQR(val,na.rm=TRUE)
  outliers <- val[(val>quantile75+1.5*iq_range)|(val<quantile25-1.5*iq_range)]
  return(length(outliers))
}
num_outliers(c(0,40,60,185)) %>% print()
```

##1.5 (Q4)
Now combine your function num_outliers() with the functions group_by() and summarise() to compute
the number of outliers for the three samples of hawk weights broken down by species.

```{r 1.5-4}
Hawks %>% group_by(Species)%>% summarise(num_outliersweight=num_outliers(Weight)) %>% print()
```

##1.6 (Q1)
Compute the covariance and correlation between the Weight and Wing of the Hawks data. You can use
the cov and cor functions

```{r 1.6-1}
cov(Hawks$Weight,Hawks$Wing,use='complete.obs') %>% print()
cor(Hawks$Weight,Hawks$Wing,use='complete.obs') %>% print()
```

#1.6 (Q2)
 Suppose that we have a pair of variables: X with values X1, · · · , Xn and Y with values Y1, · · · , Yn.
Suppose that X1, · · · , Xn and Y1, · · · , Yn have the sample covariance S and correlation R. Let a, b ∈ R be
real numbers and define a new variable X˜ with X˜
1, · · · , X˜
n defined by X˜
i = aXi + b for i = 1, 2, · · · , n. In
addition, Let c, d ∈ R be real numbers and define a new variable Y˜ with Y˜
1, · · · , Y˜
n defined by Y˜
i = cYi + d.
What is the covariance between X˜
1, · · · , X˜
n and Y˜ with Y˜
1, · · · , Y˜
n (as a function of S, a, b, c, d)? Assuming
that a ̸= 0 and c ̸= 0, what is the correlation between X˜
1, · · · , X˜
n and Y˜ with Y˜
1, · · · , Y˜
n?
Let a = 2.4, b = 7.1, c = −1, d = 3, and let X be the hawk’s weight and Y be the hawk’s Wing.

```{r 1.6-2}
X = Hawks$Weight
Y= Hawks$Wing

a = 2.4
b = 7.1
c = -1
d = 3

X2 = a*X+b
Y2 = c*Y+d

cov(X2,Y2,use='complete.obs') %>% print()
cor(X2,Y2,use='complete.obs') %>% print()

print(a*c*cov(X,Y,use='complete.obs'))
print(cor(X,Y,use='complete.obs'))
```

# 2. Random experiments, events and sample spaces, and the set theory
##2.1 (Q1)
random experiment: repeatable and has a well defined set of possible outcomes
event: a set of possible outcomes 
sample space: Set of all possible outcomes

##2.1 (Q2)
Consider a random experiment of rolling a dice twice. Give an example of what is an event in this
random experiment. Also, can you write down the sample space as a set? What is the total number of
different events in this experiment? Is the empty set considered as an event?

event: a roll of the dice
sample space: {1,2,3,4,5,6}
total events: 2 events
The empty set is considered an event

##2.2 (Q1)
Let the sets A, B, C be defined by A := {1, 2, 3}, B := {2, 4, 6}, C := {4, 5, 6}.

1:What are the unions A ∪ B and A ∪ C?
A ∪ B = {1,2,3,4,6}
A ∪ C = {1,2,3,4,5,6}

2:What are the intersections A ∩ B and A ∩ C?
A ∩ B = {2}
A ∩ C = {} empty set

3:What are the complements A\B and A\C?
A\B = {1,3}
A\C = {1,2,3}

4:Are A and B disjoint? Are A and C disjoint?
A and B not disjoint
A and C are disjoint

5:Are B and A\B disjoint?
Yes

6:Write down a partition of {1,2,3,4,5,6} consisting of two sets. Also, write down another partition of
{1,2,3,4,5,6} consisting of three sets.
{1,2}{3,4,5,6}
{1,2}{3,4}{5,6}

##2.2 (Q2)
Let Ω be a sample space. Recall that for an event A ⊆ Ω the complement Ac
:= Ω\A := {w ∈ Ω : w /∈ A}.
Take a pair of events A ⊆ Ω and B ⊆ Ω.

1:Can you give an expression for (A^c)^c without using the notion of a complement?
A^c = Ω\A
(A^c)^c = Ω\ (Ω\A) = A

2: What is Ω^c?
results in empty set

3:(Subsets) Show that if A ⊆ B, then B^c ⊆ A^c
A ⊆ B: {Ω{B{A}}}
B^c = Ω\B = {Ω{}}
A^c = Ω\A = {Ω{B{}}}
Therefore B^c ⊆ A^c

4:Show that (A ∩ B)^c = A^c ∪ B^c. Let’s suppose we have a sequence of events A1, A2, · · · , AK ⊂ Ω. Can you write out an expression for (∩k=1toK Ak)^c?

This is just difficult finished rest in notebook

# 3. Visualisation

## 3 (Q1)
Use the ggplot and geom_density() functions to create the following density plot for the three species

```{r 3-1}
Hawk_plot <- ggplot(data=Hawks,aes(x=Tail,color=Species)) + xlab("Tail Length (mm)") + geom_density(adjust=0.5) + theme_bw() + ylab("Density")
Hawk_plot %>% print()
```


## 3 (Q2)
Use the ggplot and geom_violin() functions to create the following violin plot for the three species

```{r 3-2}
Hawk_plot2 <- ggplot(data=Hawks,aes(x=Tail,y=Species,fill=Species)) + xlab("Tail Length (mm)") + geom_violin() + ylab("Density")
Hawk_plot2 %>% print()
```

## 3 (Q3)
Generate a plot similar to the following plot using the ggplot() and geom_point() functions
scatter plot of tail against weight

```{r 3.3}
Hawk_plot3 <- ggplot(data=Hawks,aes(x=Tail,y=Weight,color=Species,shape=Species)) + xlab('Tail Length (mm)') + ylab('Weight (g)') + geom_point()
Hawk_plot3 %>% print()
```

1. How many aesthetics are present within the following plot?
2. What are the glyphs within this plot?
3. What are the visual cues being used within this plot?

1: 4 aesthetics x-axis, y-axis, colour and shape
2: The glyphs are points of different shape square, circle and triangle
3: The visual cues are the colour and shape of the points

## 3 (Q4)
Generate the following plot using the ggplot(), geom_point(), geom_smooth() and facet_wrap() functions.
Weight against tail for 3 different facet plots

```{r 3.4}
Hawk_plot4 <- ggplot(data=Hawks,aes(x=Tail,y=Weight,color=Species)) + xlab('Tail Length (mm)') + ylab('Weight (g)') + geom_point(na.rm=TRUE) + geom_smooth(na.rm=TRUE) + facet_wrap(~Species,scales='free_x')
Hawk_plot4 %>% print()
```

1. What are the visual cues being used within this plot?
2. Based on the plot below, what can we say about the relationship between the weight of the hawks and
their tail lengths?

1: The visual cues are the colour, and the trend line
2: Based on the plot the weight is mostly constant no matter the tail size. The weight is different depending on the species.

## 3 (Q5)
First, compute the Weight and the Tail of the heaviest hawk in the dataset. You can use filter() and select()
function to select proper data.
Second, reuse the code that you create from Q(3), adding an arrow and an annotation to indicate the heaviest
hawk.

```{r 3.5}
heaviest <- Hawks %>% filter(Weight==max(select(Hawks,Weight),na.rm=TRUE)) 
Hawk_plot5 <- ggplot(data=Hawks,aes(x=Tail,y=Weight,color=Species,shape=Species)) + xlab('Tail Length (mm)') + ylab('Weight (g)') + geom_point(na.rm=TRUE) + geom_curve(x=175,xend=as.numeric(select(heaviest,Tail)),y=1750,yend=as.numeric(select(heaviest,Weight)),arrow=arrow(length=unit(0.5,'cm')),curvature=0.1,color='black') + geom_text(x=175,y=1730,label='Heaviest Hawk',color='black')
Hawk_plot5 %>% print()
```










